Baseline centering problem

### 2018-12-19 paper/tangotd.tex

The Newton update/middle Bellman equation is
\begin{equation}
Z\gets (1+\eta)Z-\eta Z^2+\eta \gamma ZPZ
\end{equation}
which yields (after applying to $\1_{s_1}$ on the left and $\1_{s_2}/\rho(s_2)$
on the right)
\begin{multline}
Z(s_1,s_2)\gets (1+\eta)Z(s_1,s_2)-\eta \E_{s_3\sim \rho}\,
Z(s_1,s_3)Z(s_3,s_2)\,+\\\eta \gamma \E_{s_3\sim\rho,\, s'_3\sim
P(s'_3|s_3)}\,
Z(s_1,s_3)Z(s'_3,s_2)
\end{multline}
which is perfectly samplable. We can simulate the assignment
$Z(s_1,s_2)\gets...$ via gradient descent on the parameters of
$Z(s_1,s_2)$. (Or via double sampling, which is not problematic here!?)

@@

On-policy, sampling from $Z(\cdot,s_3)$ is easy: just sample one of the
recent states with geometric rates $\gamma$. This should divide the
right part of the equation by $Z(s_1,s_3)$ (up to some $\gamma$-dependent
factor). Thus we would sample two states $s_1,s_2$ along the trajectory,
sample a state $s_3$ in between on the trajectory from $s_1$, and (with
the right $\gamma$ factors) we would get a term with $Z(s_1,s_2)$, a term
with $Z(s_3,s_2)$ and a term with $Z(s'_3,s_2)$. This looks suspiciously
like a large-scale Bellman from $s_1$... Variance could be higher since
we don't use the model $Z$ and use sampling instead. To be tested. Write
carefully because $s'_3$ should be uncorrelated from $s_2$!! So not
possible with a single
trajectory?
On-policy, another
option is to sample triplets $(s_1,s_3,s_2)$ that are some time apart on
the trajectory, e.g., starting with $s_3$ uniform along the past
trajectory, and moving back for $s_1$ and forward for $s_2$ (with factors
$\gamma$). Looks a lot like eligibility traces?

On-policy, working with $\gamma\to 1$ is quite easy: indeed $Z(s_1,s_2)$ is
a density with respect to $\mu(s_2)$ so that the asymptotics when
$\gamma\to 1$ is $Z(s_1,s_2)=\frac{1}{1-\gamma}+Z_0(s_1,s_2)$, and we can
directly learn $Z_0$ by inserting $\frac{1}{1-\gamma}+Z_0$ in the middle
Bellman equation. TODO check for simplifications.

>>>>

The Newton update/middle Bellman equation is
\begin{equation}
Z\gets (1+\eta)Z-\eta Z^2+\eta \gamma ZPZ
\end{equation}
which yields (after applying to $\1_{s_1}$ on the left and $\1_{s_2}/\rho(s_2)$
on the right)
\begin{multline}
Z(s_1,s_2)\gets (1+\eta)Z(s_1,s_2)-\eta \E_{s_3\sim \rho}\,
Z(s_1,s_3)Z(s_3,s_2)\,+\\\eta \gamma \E_{s_3\sim\rho,\, s'_3\sim
P(s'_3|s_3)}\,
Z(s_1,s_3)Z(s'_3,s_2)
\end{multline}
which is perfectly samplable. We can simulate the assignment
$Z(s_1,s_2)\gets...$ via gradient descent on the parameters of
$Z(s_1,s_2)$. \footnote{Should we adjust the size of the gradient descent
to get the right change of $Z(s_1,s_2)$? If we just use SGD, since
gradients may change from point to point, the speed may be different at
various states and we depart from Newton. This is clear when $Z$ is very
small: Newton grows exponentially while a standard gradient descent would
make small steps.} (Or via double sampling, which is not problematic here!?)

@@

On-policy, sampling from $Z(\cdot,s_3)$ is easy: just sample one of the
recent states with geometric rates $\gamma$. This should divide the
right part of the equation by $Z(s_1,s_3)$ (up to some $\gamma$-dependent
factor). Thus we would sample two states $s_1,s_2$ along the trajectory,
sample a state $s_3$ in between on the trajectory from $s_1$, and (with
the right $\gamma$ factors) we would get a term with $Z(s_1,s_2)$, a term
with $Z(s_3,s_2)$ and a term with $Z(s'_3,s_2)$. This looks suspiciously
like a large-scale Bellman from $s_1$... Variance could be higher since
we don't use the model $Z$ and use sampling instead; plus, we lose any
generalization introduced by $Z$. To be tested. Write
carefully because $s'_3$ should be uncorrelated from $s_2$!! So not
possible with a single
trajectory?
On-policy, another
option is to sample triplets $(s_1,s_3,s_2)$ that are some time apart on
the trajectory, e.g., starting with $s_3$ uniform along the past
trajectory, and moving back for $s_1$ and forward for $s_2$ (with factors
$\gamma$). Looks a lot like eligibility traces?

On-policy, working with $\gamma\to 1$ is quite easy: indeed $Z(s_1,s_2)$ is
a density with respect to $\mu(s_2)$ so that the asymptotics when
$\gamma\to 1$ is $Z(s_1,s_2)=\frac{1}{1-\gamma}+Z_0(s_1,s_2)$, and we can
directly learn $Z_0$ by inserting $\frac{1}{1-\gamma}+Z_0$ in the middle
Bellman equation. This yields (omitting the expectations)
\begin{multline}
Z_0(s_1,s_2)\gets (1+\eta)Z_0(s_1,s_2)-\eta
Z_0(s_1,s_3)Z_0(s_3,s_2)+\\\eta\gamma Z_0(s_1,s_3)Z_0(s'_3,s_2)-\eta
Z_0(s_1,s_3)+\frac{\eta}{1-\gamma}(\gamma Z_0(s'_3,s_2)-Z_0(s_3,s_2))
\end{multline}
TODO CHECK which we cannot directly apply with $\gamma=1$...
